---
title: "Class 7: Machine Learning 1"
author: "Rebecca Taheri (PID A17228385)"
format: pdf
---

## Background

Today we will begin out exploration of important machine learning methods with a focus on **clustering** and **dimensionallity reduction**.

To start testing these methods let’s make up some sample data to cluster where we knoew what the answer should be.

```{r}
hist(rnorm(3000, mean = 10))
```
 
 >Q. Can you generate 30 numbers centered at +3 and 30 numbers ar -3 taken from a normal ditribution?
 
```{r}
tmp <- c(rnorm(30, mean = 3),
rnorm(30, mean = -3) )
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```
 
## K-means Clustering

The main function in “base R” for K-means clustering is called `kmeans()`, let’s try it out:

```{r}
k <- kmeans(x, centers = 2)
k
```


> Q. What component of your kmeans result object has the cluster centers?

```{r}
k$centers
```

 >Q. What component of your kmeans result object has the cluster size (i.e. how many points are in each cluster?)
 
```{r}
k$size
```
 
> Q. What component of your kmeans result object has the cluster membership vector (i.e. the main clustering result: which points are in which cluster?)

```{r}
k$cluster
```

> Q. Plot the results of clustering (i.e. our data colored by the clustering result) along with the cluster centers.

```{r}
plot(x, col = k$cluster)
points(k$centers, col = "blue", pch = 15, cex = 2)

```

> Q. can you run `kmeans()` again and cluster x into 4 clusters and plot the results just like we did above with coloring by cluster and the cluster centers shown in blue?

```{r}
k4 <- kmeans(x, centers = 4)
k4$centers
```
```{r}
plot(x, col = k4$cluster)
points(k4$centers, col = "blue", pch = 15, cex = 2)
```

 
 > **Key-Point:** Kmeans will always return the clustering that we ask for (this is the “K” or “centers” in K-means)!
 
```{r}
k$tot.withinss
```
 
 ## Hierarchial Clustering
 
The main function of Hierarchical clustering in base R is called `hclust()`.
One of the main differences with respect to the `kmeans()` function is that you can not just
pass your input data directly to `hclust()` - it needs a “distance matrix” as input. We can get
this from lot’s of places including the `dist()` function.

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can “cut” the dendrogram or “tree” at a given height to yield our “clusters”. For this, we use the function `cutree()`.

```{r}
plot(hc)
abline(h=10, col = "red")
grps <- cutree(hc, h=10)
```
```{r}
grps
```

>Q. Plot our data `x` colored by the clusterview result from `hclust()` and `cutree()`?

```{r}
plot(x, col = grps)
```

```{r}
plot(hc)
abline(h=3.2, col = "red")
grps <- cutree(hc, h=3.2)

```

## Principal Component Analysis (PCA)

PCA is a popular dimensionality reduction technique that is widely used in bioinformatics.

### PCA of UK food data

Read data on food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

It looks like the row names are not set properly. We can fix this

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```
A better way to do this is fix the row names assignment at import time:

```{r}
read.csv(url, row.names = 1)
```
> Q1. How many rows and columns are in your new data frame named x? What R
functions could you use to answer this questions?

```{r}
ncol(x)
```
```{r}
nrow(x)
```


```{r}
dim(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do
you prefer and why? Is one approach more robust than another under certain
circumstances?

I think the second one is better because it is easier and faster.


```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
> Q3. Changing what optional argument in the above barplot() function results in
the following plot

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

If we change the argument `beside` from `T` to `F` it would make a plot where the bars are stacked on top of each other.

```{r}
library(tidyr)
# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |>
tibble::rownames_to_column("Food") |>
pivot_longer(cols = -Food,
names_to = "Country",
values_to = "Consumption")
dim(x_long)
```

```{r}
# Create grouped bar plot
library(ggplot2)
ggplot(x_long) +
aes(x = Country, y = Consumption, fill = Food) +
geom_col(position = "dodge") +
theme_bw()
```


```{r}
library(ggplot2)
ggplot(x_long) +
aes(x = Country, y = Consumption, fill = Food) +
geom_col(position = "stack") +
theme_bw()
```
> Q5. We can use the `pairs()` function to generate all pairwise plots for our countries.
Can you make sense of the following code and resulting figure? What does it mean
if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

The `pairs()` function plots every variable against every other variable. When a point is on the diagonal that means that variable is being compared to itself.

## Heatmap

We can install the **pheatmap package** with the `install.packages()` command that we used previously. Remember that we always run this in the console and not a code chunk in our quarto document.

```{r}
library(pheatmap)
pheatmap( as.matrix(x) )
```
Of all these plots, really only the `pairs()` plot was useful. This however took a bit of work to interpret and will at scale when I am looking at much bigger datasets.

> Q6. Based on the pairs and heatmap figures, which countries cluster together and
what does this suggest about their food consumption patterns? Can you easily tell
what the main differences between N. Ireland and the other countries of the UK in
terms of this data-set?

England, Wales, and Scotland cluster together, indicating similar food consumption patterns. Northern Ireland stands out in both plots, suggesting its food consumption differs from the rest of the UK.

> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)
# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
aes(x = PC1, y = PC2, label = rownames(pca$x)) +
geom_point(size = 3) +
geom_text(vjust = -0.5) +
xlim(-270, 500) +
xlab("PC1") +
ylab("PC2") +
theme_bw()
```

>Q8. Customize your plot so that the colors of the country names match the colors
in our UK and Ireland map and table at start of this document

```{r}
country_cols <- c( "England" = "red", "Wales" = "green", "Scotland" = "blue", "N.Ireland" = "orange")
ggplot(pca$x) + 
  aes(x = PC1, y = PC2, label = rownames(pca$x), color = rownames(pca$x)) + geom_point(size = 3) +
geom_text(vjust = -0.5) + scale_color_manual(values = country_cols) + 
xlim(-270, 500) +
xlab("PC1") +
ylab("PC2") +
theme_bw() + theme(legend.position = "none")
  
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
aes(x = PC2,
y = reorder(rownames(pca$rotation), PC2)) +
geom_col(fill = "steelblue") +
xlab("PC2 Loading Score") +
ylab("") +
theme_bw() +
theme(axis.text.y = element_text(size = 9))
```

The two that stand out the most for PC2 are soft drinks and fresh potatoes, PC2 shows the differences in how much these food are consumed aross countries.

>Q10: How many genes and samples are in this data set? How many PCs do you think it will take to have a useful overview of this data set (see below)?

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
There are 6 genes and 10 samples in this dataset.  I think it will take around 3 PCs to have a useful overview of this data set.


## PCA The Rescue

```{r}
pca <- prcomp(t(x))
summary(pca)
```

> Q. How much variance is captured in the first PC?

67.4%

> Q. How many PCs do I need to capture at least 90% of the total variance in the dataset?

2 PCs capture 96.5% of the total variance


> Q. Plot our main PCA result. Folks like to call this different things depending on
their field of study e.g. “PC plot”, “ordienation plot”, “Score plot”, “PC1 vs. PC2
plot”…

```{r}
attributes(pca)
```

```{r}
pca$x
```


```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col = my_cols, pch=16) 
```

```{r}
library(ggplot2)
ggplot(pca$x) +
aes(PC1, PC2) +
geom_point(col = my_cols)
```



