---
title: "Homework class 12"
author: "Rebecca Taheri ( Pid: A17228385)"
format: pdf
---

## Section 4 Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

total samples = 462
```{r}
table(expr$geno)

```

sample sizes are 108. for A/A, 233 for A/G, and 121 for G/G
```{r}
tapply(expr$exp, expr$geno, median)

```

The median expression levels for each geno type are 31.248 for A/A,  25.065 for A/G, 20.074 for G/G

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch = TRUE) +
   labs(title = "ORMDL3 Expression by Genotype",
       x = "Genotype",
       y = "Expression Level")
```

The boxplot above shows that A/A individuals have higher ORMDL3 expression than G/G individuals. This means that the G allele is linked to having lower expression levels. So I can conclude that this SNP does affect ORMDL3 expression.



