---
title: "class 5: data Viz with ggplot"
author: "Rebecca ( Pid: A17228385)"
format: gfm
toc: true
---

## Background

There are lot's of ways to make figures in R. These include so-called "base R" graphics (e.g `plot()`) and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command
`install.package()`.

> **N.B** We never run an instal cmd in quarto code chunk or we will end up re-installing packages many many times - which is not what we want

Everytime we want to use one of these "add-on" packages we need to load it up in R with `library()` fuction:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data map to the plot
- The**geom**etry, the type of plot

Add a line to better show the relationship between speed and dist

```{r}
p <- ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) + 
  labs(title = "Stopping distance of old cars",
       subtitle="Data from the `cars` object", 
       x= " Speed (mph)", 
       y = "Distance (ft)")
```

Render it out

```{r}
p + theme_bw()
```
 
## Gene expression plot

We can read the input data from the class website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot 


```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) +
  geom_point()
```

```{r}
table(genes$State)
```

Version 2 let's color by `Stat` so we can see the "up" and down significant genes compared to all the "unchanging gene"


```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Version 3 plot, let's modify the default colors to something we like


```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) +
  geom_point() + 
  scale_color_manual(values=c("Blue","grey","purple")) +
  labs( x="Control (no drug)",
        y = "Drug", 
        title = "Gene Expression Changes upon GLP-1") +
  theme_bw()
```


## Going Further

Let's have a look at the famous **gapminder** dataset

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```


```{r}
head(gapminder, 3)
```

```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3) 
```

Let's "facet ( i.e make a separate plot) by continent rather than the big hot mess above.

```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3) +
  facet_wrap(~continent)

```


## Custom plots

How big is this gapminder dataset?

```{r}
nrow(gapminder)
```
I want to "filter" down to a subset of this data. I will use **dplyr** package to help me.

First I need to install it and the load it up..
`install.packages("dplyr")` and then `library(dplyr)`

```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```


```{r}
filter(gapminder_2007, country=="Ireland")
```

```{r}
filter(gapminder, year==2007, country=="Ireland")
```

```{r}
filter(gapminder, year==2007, country=="United States")
```

> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
gapminder_c <- filter(gapminder, year %in% c(1977,2007))
```

```{r}
ggplot(gapminder_c) + aes(gdpPercap, lifeExp, col=continent) +
  geom_point() + 
  facet_wrap(~year)
```

